### 2020-07-18 paper.tex
minor proofreading corrections

\begin{abstract}
Involutive MCMC is a unifying mathematical construction for MCMC kernels that generalizes many classic and state-of-the-art MCMC algorithms, from reversible jump MCMC to kernels based on deep neural networks.
But like MCMC samplers more generally, implementing involutive MCMC kernels is often tedious and error-prone, especially when sampling on complex state spaces.
This paper describes a technique for automating the implementation of involutive MCMC kernels given (i) a pair of probabilistic programs defining the target distribution and an auxiliary distribution respectively and (ii) a differentiable program that transforms the execution traces of these probabilistic programs.
The technique, which is implemented as part of the Gen probabilistic programming system, also automatically detects user errors in the specification of involutive MCMC kernels and exploits sparsity in the kernels for improved efficiency.
The paper shows example Gen code for a split-merge reversible jump move in an infinite Gaussian mixture model and a state-dependent mixture of proposals on a combinatorial space of covariance functions for a Gaussian process.
\end{abstract}

\section{INTRODUCTION}
Markov chain Monte Carlo (MCMC) algorithms are powerful tools for approximate sampling from probability distributions and are central to modern Bayesian statistics, probabilistic machine learning, statistical physics, and numerous applications of probabilistic inference.
But designing and deriving efficient MCMC algorithms is mathematically involved, and implementing MCMC kernels is tedious and notoriously error-prone.
These challenges are especially pertinent when sampling from probability distributions on complex state spaces that combine symbolic, numeric, and structural uncertainty, such as those arising in computational biology~\citep{huelsenbeck2004bayesian}, robotics and scene understanding~\citep{geiger2011generative}, and models of human cognition~\citep{tenenbaum2011grow}.

Involutive MCMC is a mathematical construction for MCMC kernels that gives a simplifying and unifying perspective on a number of previously disparate classes of kernels,
including reversible jump MCMC~\citep{green1995reversible}, which is the dominant mathematical framework for MCMC on complex state spaces.
Involutive MCMC constructs an MCMC kernel from three components: (i) the unnormalized target density, (ii) a sampler and density for an auxiliary probability distribution (iii) an involution\footnote{a bijection that is its own inverse ($f$ where $f(f(z)) = z$)} on an extended state space.
While this construction is mathematically clarifying, correctly implementing involutive MCMC kernels on complex state spaces remains challenging due to tedious density and Jacobian computations and the need for careful reasoning about the state space.

This paper formulates involutive MCMC on general state spaces and shows how to automate the implementation of an involutive MCMC kernel from three declarative programs that define the target probability distribution, auxiliary probability distribution, and the involution, respectively.
The probability distributions are defined as probabilistic programs, and the involution is defined as a differentiable program that transforms the execution traces of the probabilistic programs.
We use probabilistic programming techniques and automatic differentiation to automatically compute the acceptance probability.
We also show how to automatically detect mathematical errors in the specification of an involutive MCMC kernel, and how to improve the efficiency by automatically exploiting the sparsity structure in the involution.
We implemented the approach within the Gen probabilistic programming system\footnote{\url{https://www.gen.dev}}~\citep{cusumano2019gen}.
The paper shows examples of involutive MCMC kernels implemented in Gen for (i) a split-merge reversible jump move in an infinite mixture model, and a (ii) state-dependent mixture of Metropolis-Hastings proposals on an infinite combinatorial space of covariance functions for a Gaussian process.
We also provide a lightweight PyTorch implementation of the basic approach at
{\small\url{https://github.com/probcomp/autoimcmc}}.

The contributions of this paper include:
\begin{enumerate}
\item A measure-theoretic formulation of involutive MCMC on general state spaces.
\item A mathematical formulation of state spaces consisting of arbitrary key-value stores (i.e. dictionaries) that formalizes the space of execution traces of Gen probabilistic programs, and a formulation of involutive MCMC on these spaces.
\item A differentiable programming language for defining bijections between spaces of traces of probabilistic programs.
\item An algorithm that automates the implementation of an involutive MCMC kernel given two probabilistic programs and a differentiable program encoding the involution, using automatic differentiation and tracing of probabilistic programs.
\item An extension to the algorithm that exploits sparsity in the involution to reduce the computational complexity of an involutive MCMC kernel from $O(N^2)$ to $O(1)$ in common cases where $N$ is the dimensionality of the latent space.
\item An algorithm that dynamically detects errors in the specification of an involutive MCMC kernel.
\end{enumerate}

@@

The involutive MCMC construction was previously implemented~\citep{gen-inv-mcmc} by the authors as part of the inference library of the Gen probabilistic programming system~\citep{cusumano2019gen}.
The construction was motivated in part by a desire for a simple interface that automated the implementation of reversible jump MCMC samplers~\citep{green1995reversible}, state-dependent mixtures of proposals on complex state spaces, and data-driven neural proposals.
Gen's involutive MCMC construction has since been used by a number of researchers to design and implement MCMC algorithms in diverse domains including computational biology~\citep{merrell2020inferring} and artificial intelligence~\citep{zhi2020online}.

\citet{imcmc} independently identified the involutive MCMC construction as a unifying framework for MCMC algorithms, and showed how a number of classic and recent MCMC algorithms can be cast within this framework.
\citet{imcmc} also identified design principles for developing new MCMC algorithms using the involutive MCMC construction, and showed that the framework aids in the derivation of novel efficient MCMC algorithms.

@@

In recent decades, a number of probabilistic programming systems have provided automated implementations of probabilistic inference algorithms~\citep{gilks1994language,milch2007,pfeffer200714,goodman2012church,gehr2016psi,carpenter2017stan}.
The majority of these systems have focused on the automation of one or more built-in inference algorithms, with customization limited tuning parameters of the algorithm.
The programmable inference paradigm~\citep{mansinghka2018probabilistic} to probabilistic programming instead emphasizes user control over the inference algorithm and automation of the lower-level implementation.
In particular, the Gen probabilistic programming system~\citep{cusumano2019gen} offers a high degree of user programmability by exposing an API that supports high-level user implementations of inference algorithms, instead of a set of a built-in inference engine.
One other probabilistic programming system besides Gen supports custom reversible jump samplers:
\citet{roberts2019} presents a system embedded in Haskell that can automatically generate the implementation of certain reversible jump MCMC kernels from a high-level specification, but does not handle many kernels that can be handled by our technique, including the example in Figure~\ref{fig:structure-learning}.

@@

Involutive MCMC is a general framework for constructing MCMC kernels that are stationary
for a target probability distribution $p$. Informally, the algorithm works as follows: starting at some state $x$, we first sample
an auxiliary variable $y \sim q_x$ from a state-dependent auxiliary distribution. We then apply an \textit{involution} $f$ to the pair $(x, y)$ to obtain $(x', y')$. 
Finally, we computes an acceptance probability $\alpha$ and either accept $x'$ as the new state, or reject it and repeat the previous state $x$.
Different choices of $q$ and $f$ recover many algorithms from the literature~\citep{imcmc}.

@@

This section describes probability distributions on dictionaries, and gives a constructive definition of the involutive MCMC acceptance probability in this setting in terms of a Jacobian.
Section~\ref{sec:automating} will then show how probability distributions on dictionaries can be specified with probabilistic programs, and how probabilistic programming techniques can automatically compute probability densities spaces of dictionaries.

@@

\textbf{Relationship to representation of~\citet{green1995reversible}.}
\citet{green1995reversible} uses a state space that is the countable union of `models', where for each model there is a vector of real-valued parameters.
Dictionaries possesses substantially more structure:
Instead of monolithic `models', dictionaries use a more elaborate discrete state that includes the set of keys and the assignment to the discrete keys.
Also, continuous keys play the role of real-valued parameters, so it is possible to express that a given real-valued parameter is shared between models.
The additional structure of dictionaries enables the automation techniques in Section~\ref{sec:automating}.

\textbf{A measure space of finite dictionaries.}
We associate a measure $\mu_k$ on $V_k$ for each key---the counting measure for each discrete key and the Lebesgue-measure on $\mathbb{R}^{d_k}$ for each continuous key.
For each finite set of keys $K$, we make $V_K$ a measure space using the standard product $\sigma$-algebra $\Sigma_K = \otimes_{k \in A} \Sigma_k$ and the product measure $\mu_K = \times_{k \in A} \mu_k$.
We equip $\mathcal{D}$ with the $\sigma$-algebra $\Sigma_{\mathcal{D}} := \{\bigcup_{K \subset \mathcal{K}, |K| < \infty} \{(K, \mathbf{x}) \mid \mathbf{x} \in B_K\} \mid B_K \in \Sigma_K \, \text{for each finite } K \subset \mathcal{K}\}$ to obtain a measurable space
of dictionaries.
A reference measure $\mu_\mathcal{D}$ on this space can be constructed using the product measures $\mu_K$: we set $\mu(B) := \sum_{A \in \mathcal{K}, |K| < \infty} \mu_K(\{\mathbf{x} \mid (K, \mathbf{x}) \in B\})$.

@@

\textbf{Probability distributions on finite dictionaries.} 
When all keys are discrete, a probability distribution on dictionaries is defined by a $p$ is a probability mass function $p : \mathcal{D} \to [0, 1]$ that assigns a probability $p(m)$ to each dictionary $m \in \mathcal{P}$ such that $\sum_{x \in \mathcal{D}} p(m) = 1$.
More generally a probability distribution on dictionaries is defined by a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$. 
The probability is distributed the finite sets of keys $K \subseteq \mathcal{K}$ where $|K| < \infty$:
\begin{equation}
1 = \int p(m) \mu_\mathcal{D}(\text{d}m) = \sum_{\substack{K \subseteq \mathcal{K}\\|K| < \infty}} \int p(m) \mu_{K}(dm)
\end{equation}
where the probability for each key set $K$ is:
\begin{equation}\label{eq:integral}
\int p(m) \mu_{K}(dm) = \sum_{\mathbf{x}_1} \int_{\mathbb{R}^{d}} p((K, (\mathbf{x}_1, \mathbf{x}_2))) d\mathbf{x}_2
\end{equation}
where $d: = \sum_{k \in A \cap \mathcal{J}} d_{k}$ and where $\mathbf{x}_1$ is an assignment to the discrete choices in $K$ and $\mathbf{x}_2$ is an assignment to the continuous choices.

We now give an example to build intuition.
Consider a generative model of univariate data points $y_1, \dots, y_n$ from a Gaussian mixture with an unknown number of components $k$, each with unknown mean $m_i$ and variance $s_i$. If we place a Gaussian prior on $m_i$, an inverse Gamma prior on $s_i$, and a Poisson prior on $k$, the resulting density on dictionaries $d$ is:
%
\begin{align*}
p(d) =
\begin{array}{l}
p_{\mathrm{poisson}(3)}(d[\mathtt{k}]) \cdot\\
\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{normal}(0, 1)}(d[\mathtt{m}_i]) \cdot\\
\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{inversegamma}(1, 10)}(d[\mathtt{s}_i]) \cdot\\
\prod_{i=1}^n  \frac{1}{d[\mathtt{k}]} \sum_{j=1}^{d[\mathtt{k}]} \cdot p_{\mathrm{normal}(d[\mathtt{m}_j], d[\mathtt{s}_i])}(d[\mathtt{y}_i])
\end{array}
\end{align*}
%
when $K_d = \{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq d[\texttt{k}]\} \cup \{\texttt{s}_i \mid 1 \leq i \leq d[\texttt{k}]\}$, and 0 otherwise. In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other keys $k \in \mathcal{K}$, $V_k = \mathbb{R}$ with the Lebesgue measure for $\mu_k$.
For each $k \in \{0, 1, \ldots\}$, the probability for key set
$\{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$
is $p_{\mathrm{poisson}(s)}(k)$ (via Equation~(\ref{eq:integral})).

@@

Suppose that the model distribution and auxiliary distributions are probability distributions on dictionaries, with densities $p$ and $q_x$.
Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{K}_P$ and $\mathcal{K}_Q$ respectively, so that $Z = X \times Y \subseteq \mathcal{D}_P \times \mathcal{D}_Q$ where $\mathcal{D}_P$ is the set of dictionaries on keys taken from $\mathcal{K}_P$ and similarly for $\mathcal{D}_Q$.
% TODO: this is only needed for Algorithm 2, and it is relevant for the 
To simplify the notation, and without loss of generality, we will assume that $\mathcal{K}_P$ and $\mathcal{K}_Q$ are disjoint\footnote{If $\mathcal{K}_P$ and $\mathcal{K}_Q$ are not disjoint then, they can be made so by adding a different prefix to the keys of each set.}, and we define $Z := \{ x \oplus y : p(x)q_x(y) > 0\} \subseteq \mathcal{D}$.
where $\mathcal{D}$ is the set of dictionaries on keys from $\mathcal{K}_P \cup \mathcal{K}_Q$
(recall $x \oplus y$ denotes the dictionary resulting from merging dictionaries $x$ and $y$ with disjoint keys).

Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $K_x$ and $K_y$ and the values $x_k$ and $y_k$ for all discrete keys $k$ are both constant.
Then each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries.
Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
Let $e(z) \in E$ denote the element of the partition for an alement $z \in Z$.
Then, $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
\begin{equation*}
f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z
\end{equation*}
Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
Then, the acceptance probability in Algorithm~\ref{alg:involutive-mcmc} simplifies to:
\begin{equation} \label{eq:acceptance-ratio-dictionaries}
\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
\end{equation}
One example of a valid partition of $Z$ is given by equivalence classes of the following equivalence relation:
\[
z_1 \sim z_2 \iff (K_{z_1} = K_{z_2}) \land (z_1[k] = z_2[k] \; \forall k \in K_{z_1} \cap \mathcal{I})
\]
(dictionaries are equivalent if they contain the same keys and they agree on the value of all discrete keys).
Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.

@@

Thus, this ratio can be computed term-by-term using \textsc{score} on the probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$: (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $\mathcal{Q}_x$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $\mathcal{L}(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $\mathcal{P}$ with trace $x \oplus b$. The density $q_x(y)$ can be computed while $y$ is being sampled, using \textsc{trace-and-score}.

Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution and resulting cancellation in the ratio for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.

@@

Section~\ref{sec:probabilistic-programming} showed that if the densities $p$ and $q$ are specified using probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$, then density ratio in the acceptance probability for involutive MCMC on dictionaries Equation~(\ref{eq:acceptance-ratio-dictionaries}) can be automated using probabilistic programming techniques.
This section shows that if the involution $f$ is specified in using a differentiable program $\mathcal{F}$ that transforms the execution traces of probabilistic programs, then the Jacobian factor in Equation~(\ref{eq:acceptance-ratio-dictionaries}) can be automated using automatic differentiation.
The procedure \textproc{auto-involutive-mcmc} in Algorithm~\ref{alg:auto-involutive-mcmc} combines these two ideas and automates involutive MCMC given the programs $\mathcal{P}$, $\mathcal{Q}$, and $\mathcal{F}$.

>>>>

\begin{abstract}
Involutive MCMC is a unifying mathematical construction for MCMC kernels that generalizes many classic and state-of-the-art MCMC algorithms, from reversible jump MCMC to kernels based on deep neural networks.
But as with MCMC samplers more generally, implementing involutive MCMC kernels is often tedious and error-prone, especially when sampling on complex state spaces.
This paper describes a technique for automating the implementation of involutive MCMC kernels given (i) a pair of probabilistic programs defining the target distribution and an auxiliary distribution respectively and (ii) a differentiable program that transforms the execution traces of these probabilistic programs.
The technique, which is implemented as part of the Gen probabilistic programming system, also automatically detects user errors in the specification of involutive MCMC kernels and exploits sparsity in the kernels for improved efficiency.
The paper shows example Gen code for a split-merge reversible jump move in an infinite Gaussian mixture model and a state-dependent mixture of proposals on a combinatorial space of covariance functions for a Gaussian process.
\end{abstract}

\section{INTRODUCTION}
Markov chain Monte Carlo (MCMC) algorithms are powerful tools for approximate sampling from probability distributions and are central to modern Bayesian statistics, probabilistic machine learning, statistical physics, and numerous application areas where probabilistic modeling and inference are used.
But designing and deriving efficient MCMC algorithms is mathematically involved, and implementing MCMC kernels is tedious and notoriously error-prone.
These challenges are especially pronounced when sampling from probability distributions on complex state spaces that combine symbolic, numeric, and structural uncertainty, such as those arising in computational biology~\citep{huelsenbeck2004bayesian}, robotics and scene understanding~\citep{geiger2011generative}, and models of human cognition~\citep{tenenbaum2011grow}.

Involutive MCMC is a mathematical construction for MCMC kernels that gives a simplifying and unifying perspective on a number of previously disparate classes of kernels,
including reversible jump MCMC~\citep{green1995reversible}, which is the dominant mathematical framework for MCMC on complex state spaces.
Involutive MCMC constructs an MCMC kernel from three components: (i) the unnormalized target density, (ii) a sampler and density for an auxiliary probability distribution, and (iii) an involution\footnote{a bijection that is its own inverse ($f$ where $f(f(z)) = z$)} on an extended state space.
While this construction is mathematically clarifying, correctly implementing involutive MCMC kernels on complex state spaces remains challenging due to tedious density and Jacobian computations and the need for careful reasoning about the state space.

This paper formulates involutive MCMC on general state spaces and shows how to automate the implementation of an involutive MCMC kernel from three declarative programs that define the target probability distribution, auxiliary probability distribution, and the involution, respectively.
The probability distributions are defined as probabilistic programs, and the involution is defined as a differentiable program that transforms the execution traces of the probabilistic programs.
We use probabilistic programming techniques and automatic differentiation to automatically compute the acceptance probability.
We also show how to automatically detect mathematical errors in the specification of an involutive MCMC kernel, and how to improve the efficiency by automatically exploiting the sparsity structure in the involution.
We implemented the approach within the Gen probabilistic programming system\footnote{\url{https://www.gen.dev}}~\citep{cusumano2019gen}.
The paper shows examples of involutive MCMC kernels implemented in Gen for (i) a split-merge reversible jump move in an infinite mixture model, and (ii) a state-dependent mixture of Metropolis-Hastings proposals on an infinite combinatorial space of covariance functions for a Gaussian process.
We also provide a lightweight PyTorch implementation of the basic approach at
{\small\url{https://github.com/probcomp/autoimcmc}}.

The contributions of this paper include:
\begin{enumerate}
\item A measure-theoretic formulation of involutive MCMC on general state spaces.
\item A mathematical formulation of state spaces consisting of arbitrary key-value stores (i.e. dictionaries) that formalizes the space of execution traces of Gen probabilistic programs, and a formulation of involutive MCMC on these spaces.
\item A differentiable programming language for defining bijections between spaces of traces of probabilistic programs.
\item An algorithm that automates the implementation of an involutive MCMC kernel given two probabilistic programs and a differentiable program encoding the involution, using automatic differentiation and tracing of probabilistic programs.
\item An extension to the algorithm that exploits sparsity in the involution to reduce the computational complexity of an involutive MCMC kernel from $O(N^2)$ to $O(1)$ in common cases, where $N$ is the dimensionality of the latent space.
\item An algorithm that dynamically detects errors in the specification of an involutive MCMC kernel.
\end{enumerate}

@@

The involutive MCMC construction was previously implemented~\citep{gen-inv-mcmc} by the authors as part of the inference library of the Gen probabilistic programming system~\citep{cusumano2019gen}.
The construction was motivated in part by a desire for a simple interface that automated the implementation of reversible jump MCMC samplers~\citep{green1995reversible}, state-dependent mixtures of proposals on complex state spaces, and data-driven neural proposals.
Gen's involutive MCMC construction has since been used by a number of researchers to design and implement MCMC algorithms in diverse domains, including computational biology~\citep{merrell2020inferring} and artificial intelligence~\citep{zhi2020online}.

\citet{imcmc} independently identified the involutive MCMC construction as a unifying framework for MCMC algorithms, and showed how more than a dozen classic and recent MCMC algorithms can be cast within this framework.
\citet{imcmc} also identified design principles for developing new MCMC algorithms using the involutive MCMC construction, and showed that the framework aids in the derivation of novel efficient MCMC algorithms.

@@

In recent decades, a number of probabilistic programming systems have provided automated implementations of probabilistic inference algorithms~\citep{gilks1994language,milch2007,pfeffer200714,goodman2012church,gehr2016psi,carpenter2017stan}.
The majority of these systems have focused on the automation of one or more built-in inference algorithms, with customization limited to tuning a handful of parameters of the algorithm.
The programmable inference paradigm~\citep{mansinghka2018probabilistic} in probabilistic programming instead emphasizes user control over the inference algorithm and automation of the lower-level implementation.
In particular, the Gen probabilistic programming system~\citep{cusumano2019gen} offers a high degree of user programmability by exposing an API that supports high-level user implementations of inference algorithms, instead of a set of a built-in inference engine.
One other probabilistic programming system besides Gen supports custom reversible jump samplers:
\citet{roberts2019} present a system embedded in Haskell that can automatically generate the implementation of certain reversible jump MCMC kernels from a high-level specification, and \citet{narayanan2020symbolic} also give a technique that can be used to automatically compute the Metropolis-Hastings acceptance probabilities in some settings; however, these approaches do not handle many kernels that can be handled by our technique, including the example in Figure~\ref{fig:structure-learning}.

@@

Involutive MCMC is a general framework for constructing MCMC kernels that are stationary
for a target probability distribution $p$. Informally, the algorithm works as follows: starting at some state $x$, we first sample
an auxiliary variable $y \sim q_x$ from a state-dependent auxiliary distribution. We then apply an \textit{involution} $f$ to the pair $(x, y)$ to obtain $(x', y')$. 
Finally, we compute an acceptance probability $\alpha$ and either accept $x'$ as the new state, or reject it and repeat the previous state $x$.
Different choices of $q$ and $f$ recover many algorithms from the literature~\citep{imcmc}.

@@

This section describes probability distributions on dictionaries, and gives a constructive definition of the involutive MCMC acceptance probability in this setting in terms of a Jacobian.
Section~\ref{sec:automating} will then show how probability distributions on dictionaries can be specified with probabilistic programs, and how probabilistic programming techniques can automatically compute probability densities on spaces of dictionaries.

@@

\textbf{Relationship to representation of~\citet{green1995reversible}.}
\citet{green1995reversible} uses a state space that is the countable union of `models', where for each model there is a vector of real-valued parameters.
Dictionaries possess substantially more structure:
Instead of monolithic `models', dictionaries use a more elaborate discrete state that includes the set of keys and the assignment to the discrete keys. 
Also, because continuous keys play the role of real-valued parameters, it is possible to express that a given real-valued parameter is shared between models.
The additional structure of dictionaries enables the automation techniques in Section~\ref{sec:automating}.

\textbf{A measure space of finite dictionaries.}
We associate a measure $\mu_k$ on $V_k$ with each key $k$---the counting measure for each discrete key and the Lebesgue-measure on $\mathbb{R}^{d_k}$ for each continuous key.
For each finite set of keys $K$, we make $V_K$ a measure space using the standard product $\sigma$-algebra $\Sigma_K = \otimes_{k \in A} \Sigma_k$ and the product measure $\mu_K = \times_{k \in A} \mu_k$.
We equip $\mathcal{D}$ with the $\sigma$-algebra $\Sigma_{\mathcal{D}} := \{\bigcup_{K \subset \mathcal{K}, |K| < \infty} \{(K, \mathbf{x}) \mid \mathbf{x} \in B_K\} \mid B_K \in \Sigma_K \, \text{for each finite } K \subset \mathcal{K}\}$ to obtain a measurable space
of dictionaries.
A reference measure $\mu_\mathcal{D}$ on this space can be constructed using the product measures $\mu_K$: we set $\mu(B) := \sum_{A \in \mathcal{K}, |K| < \infty} \mu_K(\{\mathbf{x} \mid (K, \mathbf{x}) \in B\})$.

@@

\textbf{Probability distributions on finite dictionaries.} 
When $V_k$ is discrete for all keys $k$, a probability distribution on dictionaries is defined by a probability mass function $p : \mathcal{D} \to [0, 1]$ that assigns a probability $p(m)$ to each dictionary $m \in \mathcal{P}$ such that $\sum_{m \in \mathcal{D}} p(m) = 1$.
More generally a probability distribution on dictionaries is defined by a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$. 
The probability is distributed among the finite sets of keys $K \subseteq \mathcal{K}$ where $|K| < \infty$:
\begin{equation}
1 = \int p(m) \mu_\mathcal{D}(\text{d}m) = \sum_{\substack{K \subseteq \mathcal{K}\\|K| < \infty}} \int p(m) \mu_{K}(dm)
\end{equation}
where the probability for each key set $K$ is:
\begin{equation}\label{eq:integral}
\int p(m) \mu_{K}(dm) = \sum_{\mathbf{x}_1} \int_{\mathbb{R}^{d}} p((K, (\mathbf{x}_1, \mathbf{x}_2))) d\mathbf{x}_2
\end{equation}
where $d: = \sum_{k \in A \cap \mathcal{J}} d_{k}$ and where $\mathbf{x}_1$ is an assignment to the discrete choices in $K$ and $\mathbf{x}_2$ is an assignment to the continuous choices.

We now give an example to build intuition.
Consider a generative model of univariate data points $y_1, \dots, y_n$ from a Gaussian mixture with an unknown number of components $k$, each with unknown mean $m_i$ and variance $s_i$. If we place a Gaussian prior on $m_i$, an inverse Gamma prior on $s_i$, and a Poisson prior on $k$, the resulting density on dictionaries $d$ is:
%
\begin{align*}
p(d) =
\begin{array}{l}
p_{\mathrm{poisson}(3)}(d[\mathtt{k}]) \cdot\\
\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{normal}(0, 1)}(d[\mathtt{m}_i]) \cdot\\
\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{inversegamma}(1, 10)}(d[\mathtt{s}_i]) \cdot\\
\prod_{i=1}^n  \frac{1}{d[\mathtt{k}]} \sum_{j=1}^{d[\mathtt{k}]} \cdot p_{\mathrm{normal}(d[\mathtt{m}_j], d[\mathtt{s}_i])}(d[\mathtt{y}_i])
\end{array}
\end{align*}
%
when $K_d = \{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq d[\texttt{k}]\} \cup \{\texttt{s}_i \mid 1 \leq i \leq d[\texttt{k}]\}$, and 0 otherwise. In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other keys $k \in \mathcal{K}$, $V_k = \mathbb{R}$ with the Lebesgue measure for $\mu_k$.
For each $j \in \{0, 1, \ldots\}$, the probability for key set
$\{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq j\} \cup \{\texttt{s}_i \mid 1 \leq i \leq j\}$
is $p_{\mathrm{poisson}(s)}(j)$ (via Equation~(\ref{eq:integral})).

@@

Suppose that the model distribution and auxiliary distributions are probability distributions on dictionaries, with densities $p$ and $q_x$.
Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{K}_P$ and $\mathcal{K}_Q$ respectively, so that $Z = X \times Y \subseteq \mathcal{D}_P \times \mathcal{D}_Q$ where $\mathcal{D}_P$ is the set of dictionaries on keys taken from $\mathcal{K}_P$ and similarly for $\mathcal{D}_Q$.
% TODO: this is only needed for Algorithm 2, and it is relevant for the 
To simplify the notation, and without loss of generality, we will assume that $\mathcal{K}_P$ and $\mathcal{K}_Q$ are disjoint\footnote{If $\mathcal{K}_P$ and $\mathcal{K}_Q$ are not disjoint then, they can be made so by adding a different prefix to the keys of each set.}, and we define $Z := \{ x \oplus y : p(x)q_x(y) > 0\} \subseteq \mathcal{D}$,
where $\mathcal{D}$ is the set of dictionaries on keys from $\mathcal{K}_P \cup \mathcal{K}_Q$
(recall $x \oplus y$ denotes the dictionary resulting from merging dictionaries $x$ and $y$ with disjoint keys).

Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that if $(K_1, \mathbf{x}_1)$ and $(K_2, \mathbf{x}_2)$ are two dictionaries in the same component $Z_e$, then $K_1 = K_2$ and they agree on all \textit{discrete} values: $x_{1k} = x_{2k}$ for all $k \in K_1 \cap \mathcal{I}$.
Then each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries.
Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ indexed by $e \in E$, with $h_{e} = h_{g(e)}^{-1}$.
Let $e(z) \in E$ denote which element of the partition a dictionary $z \in Z$ belongs to.
Then $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
\begin{equation*}
f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z.
\end{equation*}
Let $|J h_e|(z)$ denote the absolute value of the determinant of the Jacobian of $h_e$, evaluated at $z$.
Then, the acceptance probability in Algorithm~\ref{alg:involutive-mcmc} simplifies to:
\begin{equation} \label{eq:acceptance-ratio-dictionaries}
\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
\end{equation}
One example of a valid partition of $Z$ is given by equivalence classes of the following equivalence relation:
\[
z_1 \sim z_2 \iff (K_{z_1} = K_{z_2}) \land (z_1[k] = z_2[k] \; \forall k \in K_{z_1} \cap \mathcal{I})
\]
(dictionaries are equivalent if they contain the same keys and they agree on the value of all discrete keys).
See Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.

@@

Thus, this ratio can be computed term-by-term using \textsc{score} on the probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $\mathcal{Q}_x$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $\mathcal{L}(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $\mathcal{P}$ with trace $x \oplus b$. The density $q_x(y)$ can be computed while $y$ is being sampled, using \textsc{trace-and-score}.

Section~\ref{sec:sparsity} describes a more efficient approach that exploits sparsity in the involution (and the resulting cancellations in the acceptance ratio) for improved efficiency, but requires a more sophisticated probabilistic programming runtime system.

@@

Section~\ref{sec:probabilistic-programming} showed that if the densities $p$ and $q$ are specified using probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$, then density ratio in the acceptance probability for involutive MCMC on dictionaries Equation~(\ref{eq:acceptance-ratio-dictionaries}) can be automated using probabilistic programming techniques.
This section shows that if the involution $f$ is specified in using a differentiable program $\mathcal{F}$ that transforms the execution traces of probabilistic programs, then the Jacobian factor in Equation~(\ref{eq:acceptance-ratio-dictionaries}) can also be automated, using automatic differentiation.
The procedure \textproc{auto-involutive-mcmc} in Algorithm~\ref{alg:auto-involutive-mcmc} combines these two ideas and automates involutive MCMC given the programs $\mathcal{P}$, $\mathcal{Q}$, and $\mathcal{F}$.
